---
title: "SSDD metric evaluation and radius sensitivity"
date: today
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

```{r}
#| label: load-packages
#| message: false
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tidyr)
  library(ggplot2)
  library(yardstick)
})
```

# Data processing

Building-level metrics are computed from structure footprint polygons using the `ssdd` package. For each focal building, the package identifies neighbors within a configurable search radius and computes four raw metrics in two conceptual families.

## Structure density (SD)

The SD family treats buildings as a continuous field — counts per unit area or covered-area fraction:

- **KD_raw** (quartic kernel density): a kernel-weighted neighbor count per m², using centroid-to-centroid distances within radius $r_D$. Self is excluded; an isolated building has $\text{KD} = 0$. Formally: $$\text{KD}_i = \frac{1}{\pi r_D^2} \sum_{\substack{j \ne i \\ d_{ij} \le r_D}} w_j \cdot K\!\left(\frac{d_{ij}}{r_D}\right), \quad K(u) = (1-u^2)^2$$
- **BA_raw** (basal area fraction): the fraction of the buffer area around the focal building's footprint that is covered by building polygons (including self). The buffer is a Minkowski sum of the footprint polygon, not a simple disk. Output is dimensionless in $[0, 1]$.

## Structure separation (SS)

The SS family treats buildings as a set of neighbors, computing per-neighbor inverse-distance terms (wall-to-wall) averaged over the neighbor set within radius $r_S$:

- **DP_raw** (distance proximity): mean $1/(d + \epsilon)$ over neighbors within $r_S$, where $d$ is wall-to-wall distance and $\epsilon$ is a distance floor (default 0.5 m). Nearest neighbors dominate the average. An isolated building has $\text{DP} = 0$.
- **OP_raw** (orientation proximity): like DP but each neighbor's contribution is weighted by $g(\Delta\varphi) = \exp(-(\Delta\varphi / \sigma_\theta)^2)$, a Gaussian kernel on the angle between the longest edges of the focal and neighbor buildings. Down-weights neighbors not aligned with the focal building, capturing the geometric setup relevant for radiative transfer between parallel walls. Always $\text{OP} \le \text{DP}$.

## Tuning parameters

Metric computation is governed by two primary radius parameters:

| Parameter | Default | Affects | Role |
|-----------|--------:|---------|------|
| $r_D$ | 100 m | KD, BA | Neighborhood radius for the SD family. Larger values produce a smoother density signal. |
| $r_S$ | 50 m | DP, OP | Search radius for the SS family. Defines who counts as a "neighbor". |

# Modeling approach

Four models are compared using `ranger` probability forests (500 trees) under spatial block cross-validation (500 m blocks, $k = 10$ folds):

| Model | Features | Fitting script |
|-------|----------|----------------|
| 1. Kenny benchmark | `build_dens` + terrain | `benchmark_kenny_rf.R` |
| 2. Kenny extended | `build_dens` + `area_ext_build_2` + `distance_to_nearest_building` + terrain | `benchmark_kenny_ext_rf.R` |
| 3. SSDD 3-metric | KD, BA, DP + terrain | `benchmark_ssdd_rf.R` |
| 4. SSDD 4-metric | KD, BA, DP, OP + terrain | `baseline_ssdd_rf.R` |

: {tbl-colwidths="[20, 52, 28]"}

Terrain features (elevation, slope, aspect) are included in all models. Tree-related and wind-related covariates are excluded because they are not generally available for fires beyond the LA events. All models are evaluated on identical building samples (matched via 5 m nearest-neighbor proximity join) under four CV regimes: within-fire block CV for Eaton, Palisades, and pooled data, plus leave-one-fire-out (LOFO).

`fold_aucs_rd200_rs50.csv` is assembled from the four per-model results CSVs by `assemble_fold_aucs.R`. To regenerate everything from scratch, run each of the four fitting scripts in turn, then run the assembler.

```{r}
#| label: load-data
#| message: false
sweep_all <- read_csv("../../_data/processed/sweep/sweep_all.csv",
                       show_col_types = FALSE)
sweep_results <- read_csv("../../_data/processed/sweep_cv/sweep_results.csv",
                          show_col_types = FALSE)
fold_aucs <- read_csv("fold_aucs_rd200_rs50.csv", show_col_types = FALSE)
preds <- read_csv("predictions_rd200_rs50.csv", show_col_types = FALSE)
```

# Radius sensitivity

SSDD metrics were recomputed across a refined grid of $r_D \in \{50, 100, 150, 200, 250, 300\}$ and $r_S \in \{10, 25, 50\}$. The 4-metric RF was evaluated at each grid point using identical spatial block CV folds. The initial sweep used $r_D \in \{50, 100, 150, 200\}$ and $r_S \in \{25, 50, 75, 100\}$; the refinement pushed $r_S$ lower (adding 10, dropping 75/100) after seeing that fine-scale separation carried most of the signal, while $r_D$ was extended upward with the low end preserved so that LOFO — which prefers small neighborhoods — remained on the grid.

```{r}
#| label: sweep-heatmap
#| fig-width: 9
#| fig-height: 7
sweep_summary <- sweep_results |>
  group_by(r_D, r_S, setting) |>
  summarise(auc_mean = mean(auc, na.rm = TRUE), .groups = "drop")

ggplot(sweep_summary, aes(x = factor(r_S), y = factor(r_D), fill = auc_mean)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.3f", auc_mean)), size = 3) +
  scale_fill_viridis_c(option = "C", begin = 0.4, end = 1.0) +
  facet_wrap(~ setting) +
  labs(x = expression(r[S]), y = expression(r[D]), fill = "Mean AUC") +
  theme_minimal()

ggsave('_img/radius-heatmap.png', width = 4, height = 4, units = 'in', dpi = 400)
```

Performance is largely flat across the grid, with AUC varying by only ~0.01–0.02 within any given CV setting — well within the range of fold-level noise. Neither $r_D$ nor $r_S$ shows a strong systematic effect on predictive performance. This suggests the SSDD metrics are robust to the choice of radius within this range, which is encouraging for generalizability. We select $r_D = 200$, $r_S = 50$ as a reasonable default for subsequent comparisons.

# Raw metric distributions

The density plots below show the distribution of each SSDD metric at the selected radius ($r_D = 200$, $r_S = 50$) for destroyed vs. surviving structures, split by fire. A rightward shift in the "Destroyed" density relative to "Survived" indicates a positive association with structure loss.

```{r}
#| label: metric-distributions
#| fig-width: 9
#| fig-height: 5
ssdd_cols <- c("KD_raw", "BA_raw", "DP_raw", "OP_raw")

sweep_all |>
  filter(r_D == 200, r_S == 50, !is.na(DAMAGE)) |>
  mutate(status = if_else(DAMAGE == "Destroyed (>50%)", "Destroyed", "Survived")) |>
  pivot_longer(all_of(ssdd_cols), names_to = "metric", values_to = "value") |>
  filter(value > 0) |>
  ggplot(aes(x = value, fill = status)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  facet_grid(fire ~ metric, scales = "free") +
  labs(x = NULL, y = "Density", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")

ggsave('_img/raw-metrics.png', width = 6, height = 4, units = 'in', dpi = 400)
```

# Model comparison at optimal radius

Using the best-performing radius combination from the initial sweep ($r_D = 200$, $r_S = 50$), the four models are compared.

## AUC distributions across folds

```{r}
#| label: auc-boxplots
#| fig-width: 9
#| fig-height: 5.5
model_labels <- c(
  "1_kenny"      = "1. Kenny",
  "2_kenny_ext"  = "2. Kenny extended",
  "3_ssdd_3met"  = "3. SSDD 3-metric",
  "4_ssdd_4met"  = "4. SSDD 4-metric"
)

fold_aucs |>
  mutate(model = factor(model, levels = names(model_labels), labels = model_labels)) |>
  ggplot(aes(x = setting, y = auc, fill = model)) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.7,
               outlier.size = 1.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.55, y = 0.505, label = "chance", color = "grey40",
           size = 3, hjust = 0) +
  labs(x = NULL, y = "AUC", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")

ggsave('_img/auroc-benchmarking.png', width = 6, height = 2, units = 'in', dpi = 400)
```

## ROC curves

ROC curves are constructed from held-out predictions aggregated across all CV folds — each building appears in exactly one test fold, yielding an honest out-of-sample prediction for every observation.

```{r}
#| label: roc-curves
#| fig-width: 9
#| fig-height: 7
roc_data <- preds |>
  mutate(truth = factor(truth, levels = c(0L, 1L))) |>
  group_by(model, setting) |>
  roc_curve(truth, pred, event_level = "second") |>
  ungroup() |>
  mutate(model = factor(model, levels = names(model_labels), labels = model_labels))

ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 0.7) +
  geom_abline(linetype = "dashed", color = "grey60") +
  facet_wrap(~ setting) +
  labs(x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| label: auc-summary-table
preds |>
  mutate(truth = factor(truth, levels = c(0L, 1L))) |>
  group_by(model, setting, fold) |>
  summarise(
    auc = roc_auc_vec(truth, pred, event_level = "second"),
    .groups = "drop"
  ) |>
  group_by(model, setting) |>
  summarise(median_auc = median(auc), .groups = "drop") |>
  mutate(model = factor(model, levels = names(model_labels), labels = model_labels)) |>
  pivot_wider(names_from = model, values_from = median_auc)
```
